The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
John's larger comments

The residual noise levels after flagging can be used to validate whether the flagged data are free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, DTV interference has to be detected with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{hindson-cluster-emission-2014, mwacs-2014, murphy-mwa-exoplanets-2014}.

@@

Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}. Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites. Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies. However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

>>>>

The residual noise levels after flagging can be used to validate whether the flagged data are free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (radio frequencies (RF) 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, DTV interference has to be detected with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. FM-band RFI is noticeably worse when pointing at the southern horizon, however beyond this we do not have sufficient data to explore any correlation between pointing direction and RFI. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{hindson-cluster-emission-2014, mwacs-2014, murphy-mwa-exoplanets-2014}.

@@

Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference. An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in the Digital Video Broadcasting (DVB) band 4 (167--174~MHz), which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%. The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz. Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth. The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}. Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.

@@

After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR. When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites. Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies. However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction. Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.
